\documentclass[usegeometry,parskip=half]{scrartcl}

\usepackage[
  a4paper,
  top=25mm,
  bottom=25mm,
  left=25mm,
  right=25mm,
  footskip=10mm
]{geometry}

\usepackage[hidelinks]{hyperref}
\pagenumbering{gobble}

\usepackage{microtype}
\usepackage{fontspec}
\usepackage{unicode-math}

\setmainfont{Stix Two Text}
\setsansfont{TeX Gyre Heros}
\setmathfont{Stix Two Math}

\begin{document}
\textbf{ITC8280 Fundamentals of Cryptography --- evaluation criteria}

Your final grade is determined as follows:
\begin{table}[h]
  \centering
  \begin{tabular}{ccc}
    40\% - homework assignments & 10\% - Cryptohack & 50\% - final exam
  \end{tabular}
\end{table}

To pass (i.e. get a non-zero grade), you must satisfy the following:
\begin{enumerate}
  \item You must cumulatively score at least 50\% (20/40) on your homework assignments.
  \begin{itemize}
    \item There will be at most $5$ homework assignments.
		You will have until the end (Sunday at 23:59 local time) of the 14\textsuperscript{th} week to submit your homework.
    \item There will be no extensions, except potentially for the last homework assignment.
    Each assignment that is not submitted by the final deadline is worth 0 points.
    \item Grading and grace clause: each assignment will have a soft deadline that is set at least 2 weeks after the assignment publication date.
    \begin{itemize}
      \item If you submit your assignment by then, I will review it within two weeks.
      If I find mistakes, you will have one week to fix them to gain back up to half of the lost points.
      E.g. for a homework originally graded 7/10, you can fix it to get up to 8.5/10.
      \item For assignments submitted after the soft deadline, I might review them during the semester, or wait until the final deadline.
      No grace clause is possible after the soft deadline.
    \end{itemize}
    \item When you are stuck:
    \begin{itemize}
      \item You may turn to me for help after classes/by email/by Discord.
      \item You may give pointers to each other, but not share solutions.
    \end{itemize}
  \end{itemize}
  \item {Attend the midterm (Moodle exam)}
  \begin{itemize}
    \item The midterm is mandatory, but does not affect your final grade.
		Consider it a `reality check' for your knowledge.
    \item Attendance $=$ getting at least one point on the midterm.
  \end{itemize}
  \item Attend and pass the in-person exam.
  \begin{itemize}
    \item 2h closed book written exam with a handwritten one-sided A4 cheat-sheet allowed.
    \item Alternatively: 30min oral exam (max 10 students), no cheat-sheet allowed.
    \item You must pass \emph{each} of the 6 sections in the exam ($\ge$ 3 points per section).
    \item You must get at least 1/3 of the exam points to pass the exam.
    \item 1 retake allowed.
    \item All previous requirements must be satisfied to attend.
  \end{itemize}
\end{enumerate}

The reasons for the above criteria and for the closed book exam are explained in the intro lecture.

\vspace*{\parskip}

Violation of good academic practice:

Copying someone's homework, plagiarism, non-approved AI use, etc.\footnotemark{}, fails you the course and I will report you.

\footnotetext{Chapter 9 of TalTech's academic policies: \url{https://taltech.ee/en/academic-policies-taltech}}
\end{document}
